Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-01-03 paper/paper.tex
ISSTA format + pass on text

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
%\usepackage[margin=1in]{geometry}% http://ctan.org/pkg/geometry
\usepackage{lipsum}% http://ctan.org/pkg/lipsum
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{graphicx}
\usepackage[compatibility=false]{caption}% http://ctan.org/pkg/caption
\usepackage{booktabs}
\usepackage{cite}
\usepackage{url}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{tikz}
\usetikzlibrary{matrix,fit,shapes,calc,positioning,shadows,arrows,shapes,backgrounds,decorations.markings,fadings}
\usepackage{listings}
\usepackage[caption=false, font=footnotesize]{subfig}
%%%%%%%%%%%%% code listing
\renewcommand{\ttdefault}{pcr}
\lstset{
  basicstyle=\scriptsize\ttfamily,
  keywordstyle=\scriptsize\ttfamily\bfseries,
  language=C,             % choose the language of the code
  frame=single,              % adds a frame around the code
  aboveskip=0pt,
  belowskip=0pt,
  breaklines=true,           % sets automatic line breaking
  breakatwhitespace=false,   % sets if automatic breaks should only happen at
  showspaces=false,
  %numbersep=5pt,              % Abstand der Nummern zum Text
  %tabsize=2,                  % Groesse von Tabs
  %extendedchars=true,         %
  %breaklines=true,            % Zeilen werden Umgebrochen
  keywords=[2]{tcp, flag, threshold, track, count, seconds, classtype, sid}
}
\usepackage{balance}
\usepackage{wrapfig}
\usepackage{enumitem}
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}

\input{macros}

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\begin{document}
%
\title{Synthesis of Rules for Network Intrusion Detectors}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Lucas Alcantara\MyComment{\inst{1}}\MyComment{\orcidID{0000-1111-2222-3333}} \and
Guilherme Padilha\MyComment{\inst{1}}\MyComment{\orcidID{1111-2222-3333-4444}} \and
Marcelo d'Amorim\MyComment{\orcidID{0000-0002-1323-8769}}}
%
\authorrunning{Alcantara et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Federal University of Pernambuco (UFPE), Recife, Brazil\\
\email{\{lama,ghps,damorim\}@cin.ufpe.br}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Network Intrusion Detection Systems (\nids{}) are a popular mechanism
used by system administrators to defend against network attacks. These
systems monitor the network traffic and flag suspicious network
behavior. Signature-based \nids\ do that by checking the network
traffic against a pre-defined set of rules, which can become obsolete
as attackers learn new strategies to circumvent existing defenses.
This paper proposes \tname{}\footnote{Abbreviation of \textbf{Sy}nthesis of
  Su\textbf{ri}cata R\textbf{u}le\textbf{s}.}, a technique that automatically
synthesizes \nids\ rules from positive and negative examples (\ie{},
malicious and benign traffic). \tname{} formulates synthesis as an
optimization problem whose candidate solutions are rules that
maximizes the captured the positive traffic and minimizes the captured
negative traffic. \tname{} bootstraps the search with candidate
solutions that use data from the payload of the malicious message and
produces optimal rule candidates on output. We evaluated \tname{} on a
diverse set of attacks. Results indica te that \Fix{...}

%% optimizes the
%% candidate solution to capture the positive example and, and then
%% optimizes the candidate solution to avoid capturing the negative
%% traffic.

\keywords{\nids, synthesis, search}
\end{abstract}
%
%
%
\section{Introduction}

Network Intrusion Detection Systems (\nids{}) are software systems
that monitor the network traffic for malicious behavior and act
accordingly by blocking messages or alerting humans about suspicious
events~\cite{Mitchell:2014:SID:2597757.2542049}. \nids{} are typically
placed behind a firewall, vetting the traffic that the firewall did
not block. Various open-source (\eg{}, Snort~\cite{snort} and
Suricata~\cite{suricata}) and commercial \nids implementations (\eg{},
SolarWinds~\cite{solarwinds} and IBM QRadar~\cite{qradar}) exist
today. These systems are very popular in industry to secure local
computer networks given the amount of potential malicious traffic that
exist on the Internet.

\sloppy \nids{} are typically categorized in two
groups~\cite{kumar2007survey}: rule-based and anomaly-based \nids. A
rule-based intrusion detector\footnote{\aka\ signature-based intrusion
  detector.} checks if the network traffic matches a fixed set of
rules. Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{}. This rule
prescribes a method to capture a denial-of-service
attack~\cite{understanding-dos} to a server by matching specific
conditions about the network traffic. Relevant properties about the
traffic of interest appear in bold (see
Section~\ref{sec:suri-metas-coverage}). Rule-based \nids{} focus on
known attacks whereas anomaly-based \nids{} focus on unknown attacks
that could be observed by first learning the regular traffic and then
acting
properly~\cite{kumar2007survey,Mitchell:2014:SID:2597757.2542049,cordy-etal-issta19}. These
two kinds of intrusion detection mechanisms are
complementary. Rule-based \nids could miss attacks (as rulesets can
become outdated) whereas anomaly-based \nids can raise false alarms (as
detection algorithms are approximate).

Rule-based \nids are restricted to a fixed set of rules defined by the
network system administrator. Manually writing these rules is tedious,
error-prone, but important as attackers constantly create new
strategies to circumvent existing rules. IT-security companies
capitalize on this phenomena and put rulesets on the market for
sale~\cite{proofpoint-etpro,snort-rule-subscriptions}.  Despite this
fundamental issue, rule-based systems are tremendously popular in
industry today. Anomaly-based \nids are not restricted to a fixed set
of rules. They learn usage behavior from positive traffic and, based
on that, alert uncommon behavior~\cite{7579764}. Unfortunately,
anomaly-based \nids are fundamentally imprecise. To sum up, rule-based
\nids are precise but incomplete whereas anomaly-based \nids are (more)
complete but imprecise.

%% This paper focuses on the
%% problem of discovering new rules by using anomaly-based \nids.

This paper proposes \tname{}, a technique that uses machine
intelligence to automatically synthesize rules for rule-based
\nids. Our general goal is to harden the protection of \nids by
improving the creation process of rules. One scenario of application
of \tname{} is one where an anomaly-based \nids identifies abnormal
traffic and signals that traffic to \tname{} to create and distribute
the rules.

@@

\tname{} synthesizes rules from positive and negative examples, \ie{},
malicious and benign traffic.

@@

\section{Background}

This paper focuses on signature-based \nids. Although we focused on
the Suricata \nids, the principles we used are general to any other
signature-based \nids~(\eg{}, Snort~\cite{snort}).

@@

\begin{wrapfigure}[14]{r}{0.5\textwidth}
  \centering
  \vspace{-7ex}
  \scalebox{0.7}{
    \begin{tikzpicture}
      \begin{axis}[
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          %          ylabel={\#rules},
          symbolic x coords={dns, ftp, http, icmp, ip, smb, smtp, ssh,
          tcp, tls, udp},
          xtick=data,
          xticklabel style={rotate=45},          
          nodes near coords,
          every node near coord/.append style={font=\scriptsize},        
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(dns,10) (ftp,11) (http,20) (icmp,18)(ip,16) (smb,13) (smtp,11) (ssh, 10) (tcp, 32) (tls,18) (udp, 18)};
%        \legend{relevant, irrelevant}
      \end{axis}
    \end{tikzpicture}
  }
  \vspace{-2ex}
  \caption{\label{fig:distribution-rules-protocol}Number of relevant
    options per protocol supported by Suricata.}
\end{wrapfigure}
Suricata supports a total of \totoptions\ distinct options covering
\numproto\ distinct protocols. A total of 9 of these options are for
documentation.\MyComment{For instance, the purpose of rule option \CodeIn{msg} is to print on
the output an informative message indicating what kind of intrusion
has been observed.} Figure~\ref{fig:distribution-rules-protocol} shows
the distributions of these options per protocol. The sum of
the number of protocol options is higher than \totoptions\ as some
options is shared across different protocols.

@@

\begin{table}[t!]
  \caption{\label{table:rules}Example options of Suricata.}  
  \centering
  \begin{tabular}{p{0.35cm}ll}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & controls the rule’s alert frequency\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsection{Step 1: Reverse Engineering}
\pgfplotsset{width=6cm,compat=1.8}
\pgfplotsset{every tick label/.append style={font=\tiny}}
The goal of the first step is to extract options from the positive
traffic. For that, \tname{} parses the protocol message, looking for
fields in the message associated with options (see
Section~\ref{sec:rules-and-packets}).
Unfortunately, inferring options at the field granularity
level is insufficient to capture \CodeIn{content} options (see
Section~\ref{sec:example-suricata-rules}), which refer to parts of the
payload of the message and cannot be directly extracted directly from
the fields. Handling this case is critically important as most rules
contain \CodeIn{content}
options. Figure~\ref{fig:distribution-contents} shows the histogram of
number of these options per rule (up to 10) obtained from a public
ruleset~\cite{emerging-threats-open} containing 27.8K Suricata
rules. Rules without \CodeIn{content} options is an exception and half
of the rules contain at least two \CodeIn{content} options.  To find
\begin{wrapfigure}[15]{r}{0.56\textwidth}
  \centering
  \vspace{-4ex}
  \scalebox{1.2}{  
    \begin{tikzpicture}
      \begin{axis}[
          bar width=6pt,
          scaled ticks=false,
          tick label style={/pgf/number format/fixed},
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          symbolic x coords={0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10},
          xtick=data,
          nodes near coords,
          every node near coord/.append style={font=\tiny},
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(0,25) (1,10222) (2,4956) (3,5589)
          (4,1651) (5,1457) (6, 729) (7, 471) (8, 313) (9, 137) (10, 126)};
      \end{axis}
    \end{tikzpicture}
  }
  \caption{\label{fig:distribution-contents}Histogram of number of
    \CodeIn{content} options per rule (up to 10).}
\end{wrapfigure}
the values associated with these options, \tname{} splits the payload
of the malicious message in tokens using natural language delimiters,
such as $\backslash$t, $\backslash$n, $\backslash$r, and spaces. The
rationale for this decision is that we observed, by inspecting
existing rules, that \CodeIn{content} options typically refer to
sequence of characters in the payload of messages separated by these
delimiters. This tokenization step can generate dozens of options. For
the example from Section~\ref{sec:content-example}, \tname{} produces
a total of \Fix{xx} content options.

@@

\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}%
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algnewcommand\And{\textbf{and} }
\begin{algorithm}[t!]
\textbf{INPUT:} overconstrained rule $or$ \\
\textbf{OUTPUT:} Set of minimal rules $\mathcal R$\\
\vspace{-3ex}
\captionof{algorithm}{Minimization algorithm}\label{algo1}
\begin{algorithmic}[1]
  \State $\mathcal R \gets \{\mathit{or}\}$\Comment{initial population
    of candidate solutions contains only the input rule.}\label{line-init}
  \Do\label{start-do-while}
      \State $\mathcal R' \gets \emptyset$\Comment{starting new
        iteration. new generation should be empty.}\label{init-rprime}
      \ForEach {$ro \in \mathcal R $}\Comment{analyze each candidate individual/rule separately.}\label{start-foreach}
           \For {$i \gets 1$ to $N$}\Comment{mutate each rule $N$ times.}\label{start-for}
               \State $r \gets copy(ro)$      
               \State $j\gets \mathit{rand}(0,\mathit{len}(r))$
               \State $r[j]\gets 0$\Comment{discard option $j$ of rule
                 $r$, a copy of the original rule $\mathit{ro}$}
               \If{$\mathit{isnew}(r)$ \And $\mathit{isfit}(r)$}
                   \State $\mathcal R' \cup\gets \{r\}$\Comment{$r$
                     can lead to a plausible candidate solution.}\label{line-progress}
               \EndIf
           \EndFor\label{end-for}
       \EndFor\label{end-foreach}
       \State $\mathcal R \gets R'$\Comment{update new generation}
  \doWhile{$\mathcal R'\not= \{\}$}\Comment{repeat until reaching a fixpoint}\label{end-do-while}
\end{algorithmic}
\end{algorithm}

@@

\section*{Acknowledgments}
This work is partially supported by RNP grant number 002951.

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{references}
%

\end{document}

>>>>

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,anonymous]{acmart}

\usepackage{libertine}
\usepackage{lipsum}% http://ctan.org/pkg/lipsum
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{graphicx}
\usepackage[compatibility=false]{caption}% http://ctan.org/pkg/caption
\usepackage{booktabs}
%% \usepackage{cite}
\usepackage{url}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{tikz}
\usetikzlibrary{matrix,fit,shapes,calc,positioning,shadows,arrows,shapes,backgrounds,decorations.markings,fadings}
\usepackage{listings}
\usepackage[caption=false, font=footnotesize]{subfig}
\renewcommand{\ttdefault}{pcr}
\lstset{
  basicstyle=\scriptsize\ttfamily,
  keywordstyle=\scriptsize\ttfamily\bfseries,
  language=C,             % choose the language of the code
  frame=single,              % adds a frame around the code
  aboveskip=0pt,
  belowskip=0pt,
  breaklines=true,           % sets automatic line breaking
  breakatwhitespace=false,   % sets if automatic breaks should only happen at
  showspaces=false,
  %numbersep=5pt,              % Abstand der Nummern zum Text
  %tabsize=2,                  % Groesse von Tabs
  %extendedchars=true,         %
  %breaklines=true,            % Zeilen werden Umgebrochen
  keywords=[2]{tcp, flag, threshold, track, count, seconds, classtype, sid}
}
\usepackage{balance}
\usepackage{wrapfig}
\usepackage{enumitem}
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}

%% 
\input{macros}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{10.1145/1122445.1122456}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Learning to Synthesize Rules for Network Intrusion Detectors}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Ben Trovato}
\authornote{Both authors contributed equally to this research.}
\email{trovato@corporation.com}
\orcid{1234-5678-9012}
\author{G.K.M. Tobin}
\authornotemark[1]
\email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Institute for Clarity in Documentation}
  \streetaddress{P.O. Box 1212}
  \city{Dublin}
  \state{Ohio}
  \postcode{43017-6221}
}

\author{Lars Th{\o}rv{\"a}ld}
\affiliation{%
  \institution{The Th{\o}rv{\"a}ld Group}
  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
  \city{Hekla}
  \country{Iceland}}
\email{larst@affiliation.org}

\author{Valerie B\'eranger}
\affiliation{%
  \institution{Inria Paris-Rocquencourt}
  \city{Rocquencourt}
  \country{France}
}

\author{Aparna Patel}
\affiliation{%
 \institution{Rajiv Gandhi University}
 \streetaddress{Rono-Hills}
 \city{Doimukh}
 \state{Arunachal Pradesh}
 \country{India}}

\author{Huifen Chan}
\affiliation{%
  \institution{Tsinghua University}
  \streetaddress{30 Shuangqing Rd}
  \city{Haidian Qu}
  \state{Beijing Shi}
  \country{China}}

\author{Charles Palmer}
\affiliation{%
  \institution{Palmer Research Laboratories}
  \streetaddress{8600 Datapoint Drive}
  \city{San Antonio}
  \state{Texas}
  \postcode{78229}}
\email{cpalmer@prl.com}

\author{John Smith}
\affiliation{\institution{The Th{\o}rv{\"a}ld Group}}
\email{jsmith@affiliation.org}

\author{Julius P. Kumquat}
\affiliation{\institution{The Kumquat Consortium}}
\email{jpkumquat@consortium.net}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Network Intrusion Detection Systems (\nids{}) are a popular mechanism
used by system administrators to defend against network attacks. These
systems monitor the network traffic and flag suspicious network
behavior. Signature-based \nids\ do that by checking the network
traffic against a pre-defined set of rules, which can become obsolete
as attackers learn new strategies to circumvent existing defenses.
This paper proposes \tname{}, a technique that automatically
synthesizes \nids\ rules from positive and negative examples, \ie{},
malicious and benign traffic. \tname{} formulates synthesis as an
optimization problem whose candidate solutions are rules that maximize
the capture of positive traffic and minimize the capture of negative
traffic. \tname{} bootstraps the search with candidate solutions that
use data from the payload of the malicious message and produces
optimal rule candidates on output. We evaluated \tname{} on a diverse
set of attacks. Results indicate that \Fix{...}
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{\nids, synthesis, search}

%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
\label{sec:intro}

Network Intrusion Detection Systems (\nids{}) are software systems
that monitor the network traffic for malicious behavior and act
accordingly by blocking messages or alerting humans about suspicious
events~\cite{Mitchell:2014:SID:2597757.2542049}. \nids{} are typically
placed behind a firewall, vetting the traffic that the firewall did
not block. Various open-source (\eg{}, Snort~\cite{snort} and
Suricata~\cite{suricata}) and commercial \nids\ implementations (\eg{},
SolarWinds~\cite{solarwinds} and IBM QRadar~\cite{qradar}) exist
today. These systems are very popular in industry to secure local
computer networks given the amount of potential malicious traffic that
exist on the Internet.

This paper focuses on rule-based \nids{}, which is a very popular kind
of NIDS used in industry (see Section~\ref{sec:background}). A
\emph{rule-based intrusion detector}\footnote{\aka\ signature-based
  intrusion detector.} checks if the network traffic matches a fixed
set of rules. Figure~\ref{fig:synflood-example} shows an example rule
of Suricata~\cite{suricata}, a popular open-source \nids{}. This
particular rule prescribes a method to capture a denial-of-service
attack to a server by matching specific conditions about the network
traffic~\cite{understanding-dos}. Relevant properties about the
traffic of interest appear in bold in this rule (see
Section~\ref{sec:suri-metas-coverage}). Deployments of rule-based
\nids\ are restricted to a fixed set of rules defined by the network
system administrator. Updating these rules is tedious and error-prone,
but it is important as attackers constantly create new strategies to
circumvent existing rules. IT-security companies capitalize on this
phenomena and offer rulesets on the
market~\cite{proofpoint-etpro,snort-rule-subscriptions}.

This paper proposes \tname{}\footnote{Abbreviation for
  \textbf{Sy}nthesis of Su\textbf{ri}cata R\textbf{u}le\textbf{s}.}, a
technique that uses machine intelligence to automatically synthesize
rules for rule-based \nids. Our goal is to facilitate the creation
process of rules and, consequently, harden the protection of \nids\ to
network attacks. \tname{} synthesizes rules from malicious and benign
traffic and from examples of correct rules.

@@

@@

\section{Network Intrusion Detection Systems (\nids)}
\label{sec:background}

\subsection{Rule-based and Anomaly-based \nids}

\sloppy \nids{} are typically categorized in two
groups~\cite{kumar2007survey}: rule-based (see
Section~\ref{sec:intro}) and anomaly-based \nids. An
\emph{anomaly-based intrusion detector} learn usage behavior from
benign traffic and, based on that, alert uncommon
behavior~\cite{7579764,kumar2007survey,Mitchell:2014:SID:2597757.2542049,cordy-etal-issta19}. More
precisely, an anomaly-based \nids{} flags an attack when the network
traffic manifests different characteristics compared to those of the
benign traffic.

Rule-based \nids\ can miss attacks as rulesets can become outdated
whereas anomaly-based \nids\ can report false alarms as learning
algorithms that generalize observations from bening traffic are
approximate. To sum up, rule-based \nids{} and anomaly-based \nids{}
are complementary---a rule-based \nids{} focuses on known attacks
whereas anomaly-based \nids{} focuses on unknown potential attacks.
This paper focuses on rule-based \nids\ for its tremendous popularity
in industry. Although we focused on the Suricata \nids, the principles
we used are general to any other signature-based \nids~(\eg{},
Snort~\cite{snort}).

\Mar{estou aqui-----------------}

@@

%\begin{wrapfigure}[14]{r}{0.5\textwidth}
\begin{figure}[t!]
  \centering
%  \vspace{-7ex}
  \scalebox{0.85}{
    \begin{tikzpicture}
      \begin{axis}[
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          %          ylabel={\#rules},
          symbolic x coords={dns, ftp, http, icmp, ip, smb, smtp, ssh,
          tcp, tls, udp},
          xtick=data,
          xticklabel style={rotate=45},          
          nodes near coords,
          every node near coord/.append style={font=\scriptsize},        
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(dns,10) (ftp,11) (http,20) (icmp,18)(ip,16) (smb,13) (smtp,11) (ssh, 10) (tcp, 32) (tls,18) (udp, 18)};
%        \legend{relevant, irrelevant}
      \end{axis}
    \end{tikzpicture}
  }
  \vspace{-2ex}
  \caption{\label{fig:distribution-rules-protocol}Number of relevant
    options per protocol supported by Suricata.}
%\end{wrapfigure}
\end{figure}

Suricata supports a total of \totoptions\ distinct options covering
\numproto\ distinct protocols. A total of 9 of these options are for
documentation.\MyComment{For instance, the purpose of rule option \CodeIn{msg} is to print on
the output an informative message indicating what kind of intrusion
has been observed.} Figure~\ref{fig:distribution-rules-protocol} shows
the distributions of these options per protocol. The sum of
the number of protocol options is higher than \totoptions\ as some
options is shared across different protocols.

@@

\begin{table}[t!]
  \caption{\label{table:rules}Example options of Suricata.}  
  \centering
  \begin{tabular}{clp{5.5cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & controls the rule’s alert frequency\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsection{Step 1: Reverse Engineering}

\pgfplotsset{width=6cm,compat=1.8}
\pgfplotsset{every tick label/.append style={font=\tiny}}

%\begin{wrapfigure}[15]{r}{0.56\textwidth}
\begin{figure}[t!]
  \centering
  \vspace{-4ex}
  \scalebox{1.2}{  
    \begin{tikzpicture}
      \begin{axis}[
          bar width=6pt,
          scaled ticks=false,
          tick label style={/pgf/number format/fixed},
          ybar stacked,
          enlargelimits=0.15,
          legend style={at={(0.5,-0.15)},
            anchor=north,legend columns=-1},
          symbolic x coords={0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10},
          xtick=data,
          nodes near coords,
          every node near coord/.append style={font=\tiny},
          nodes near coords align={vertical},
        ]
        \addplot coordinates {(0,25) (1,10222) (2,4956) (3,5589)
          (4,1651) (5,1457) (6, 729) (7, 471) (8, 313) (9, 137) (10, 126)};
      \end{axis}
    \end{tikzpicture}
  }
  \caption{\label{fig:distribution-contents}Histogram of number of
    \CodeIn{content} options per rule (up to 10).}
%\end{wrapfigure}
\end{figure}

The goal of the first step is to extract options from the positive
traffic. For that, \tname{} parses the protocol message, looking for
fields in the message associated with options (see
Section~\ref{sec:rules-and-packets}).
Unfortunately, inferring options at the field granularity
level is insufficient to capture \CodeIn{content} options (see
Section~\ref{sec:example-suricata-rules}), which refer to parts of the
payload of the message and cannot be directly extracted directly from
the fields. Handling this case is critically important as most rules
contain \CodeIn{content}
options. Figure~\ref{fig:distribution-contents} shows the histogram of
number of these options per rule (up to 10) obtained from a public
ruleset~\cite{emerging-threats-open} containing 27.8K Suricata
rules. Rules without \CodeIn{content} options is an exception and half
of the rules contain at least two \CodeIn{content} options.  To find
the values associated with these options, \tname{} splits the payload
of the malicious message in tokens using natural language delimiters,
such as $\backslash$t, $\backslash$n, $\backslash$r, and spaces. The
rationale for this decision is that we observed, by inspecting
existing rules, that \CodeIn{content} options typically refer to
sequence of characters in the payload of messages separated by these
delimiters. This tokenization step can generate dozens of options. For
the example from Section~\ref{sec:content-example}, \tname{} produces
a total of \Fix{xx} content options.

@@

\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}%
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algnewcommand\And{\textbf{and} }
\begin{algorithm}[t!]
  \noindent\hspace{-31ex}
  \textbf{INPUT:} overconstrained rule $or$ \\
  \noindent\hspace{-30ex} 
  \textbf{OUTPUT:} Set of minimal rules $\mathcal R$\\
%\vspace{-3ex}
\captionof{algorithm}{Minimization algorithm}\label{algo1}
\begin{algorithmic}[1]
  \State $\mathcal R \gets \{\mathit{or}\}$\Comment{initial population
    of candidate solutions contains only the input rule.}\label{line-init}
  \Do\label{start-do-while}
      \State $\mathcal R' \gets \emptyset$ \Comment{starting new
        iteration. new generation should be empty.}\label{init-rprime}
      \ForEach {$ro \in \mathcal R $}\Comment{analyze each candidate individual/rule separately.}\label{start-foreach}
           \For {$i \gets 1$ to $N$}\Comment{mutate each rule $N$ times.}\label{start-for}
               \State $r \gets copy(ro)$      
               \State $j\gets \mathit{rand}(0,\mathit{len}(r))$
               \State $r[j]\gets 0$\Comment{discard option $j$ of rule
                 $r$, a copy of the original rule $\mathit{ro}$}
               \If{$\mathit{isnew(r)} \And \mathit{isfit}(r)$} 
                   \State $\mathcal R' \cup\gets \{r\}$\Comment{$r$
                     can lead to a plausible candidate solution.}\label{line-progress}
               \EndIf
           \EndFor\label{end-for}
       \EndFor\label{end-foreach}
       \State $\mathcal R \gets R'$\Comment{update new generation}
  \doWhile{$\mathcal R'\not= \{\}$}\Comment{repeat until reaching a fixpoint}\label{end-do-while}
\end{algorithmic}
\end{algorithm}

@@

%% \section*{Acknowledgments}
%% This work is partially supported by RNP grant number 002951.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
\endinput
%%
%% End of file `sample-sigconf.tex'.
